A note on the instability and degeneracy of deep learning models

### 2016-12-01 writing/note.Rmd
nordman and vardeman final edits

---
title: "A note on the instability and degeneracy of deep learning models"
author: | 
  | Andee Kaplan, Daniel Nordman, and Stephen Vardeman
  | Iowa State University
  | ajkaplan@iastate.edu, dnordman@iastate.edu, vardeman@iastate.edu
date: "November, 2016"
linestretch: 2
output: 
  pdf_document:
    keep_tex: true
    fig_caption: yes
    includes:
      in_header: latex/header.tex
    number_sections: yes
bibliography: references/refs.bib
abstract: |
  A probability model exhibits instability if small changes in a data outcome result in large, and often unanticipated, changes in probability. For correlated data structures found in several application areas, there is increasing interest in predicting/identifying instability. We consider the problem of quantifying instability for general probability models defined on sequences of observations, where each sequence of length $N$ has a finite number of possible outcomes. (A sequence of probability models results indexed by $N$ that accommodates data of expanding dimension.) Model instability is formally shown to occur when a certain log-probability ratio under such models grows faster than $N$. In this case, a one component change in the data sequence can provably shift probability by orders of magnitude. Also, as a measure of instability becomes more extreme, the resulting probability models are shown  to tend to degeneracy, placing all their probability on arbitrarily small portions of the sample space. These results on instability apply to large classes of models commonly used in random graphs, network analysis, and machine learning contexts.
---

@@

Section \ref{examples} provides several examples of FSFS models commonly used in graph/network analysis and machine learning (i.e., deep learning models). Section \ref{instability-results} establishes formal results regarding the propriety of FSFS models with regard to stability. A FSFS probability model sequence exhibits instability if small changes in the components of a data outcome $(x_1,\ldots,x_N)$ can result in large changes in probability $P_{\theta_N}(x_1,\ldots,x_N)$. The concept of instability, introduced in the field of statistical physics by @ruelle1999statistical, was extended to include a notion of detection and quantification for certain exponential 
family models by @schweinberger2011instability. For similar exponential models, particularly in connection to random graphs/networks, @handcock2003assessing considered (mean-based) characterizations for so-called model degeneracy, whereby a probability model places all mass on a small subset of the sample space and produces undesirably low variability in model outcomes. As described by @schweinberger2011instability, model instability and model degeneracy are related by viewing degeneracy as an extreme or limiting form of instability. The instability results of @schweinberger2011instability were developed for the case of discrete exponential family models. The main results here concern a general measure of model instability, appropriate across the whole FSFS model class. This can be used to identify when certain maximal probabilities in FSFS models are too extreme relative to the length $N$. In this case, a one component change in the data sequence may provably shift probability by orders of magnitude, and FSFS models are rigorously shown to become degenerate as the measure of instability increases. Lastly, Section \ref{implications} emphasizes the implications of our model propriety results and proofs of the main results appear in the Appendix.

@@

Defining $P_{\boldsymbol \theta_N}(\boldsymbol x)\equiv p_{N,\boldsymbol \lambda_N}(\boldsymbol x)$ with $\boldsymbol \theta_N=\boldsymbol \lambda_N$ as $\boldsymbol \theta_N=\lambda$ to a sequence of elements of $\Lambda \subset \mathbb{R}^k$ and noting that $P_{\boldsymbol \theta_N}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$, these discrete exponential family models are special cases of the FSFS models. Such exponential models arise with spatial data on a lattice [@besag1974spatial], network data [@wasserman1994social; @handcock2003assessing], and even binomial sampling with $N$ iid Bernoulli random variables. (Note for random graphs or networks with, say, $m$ nodes, one may wish to consider $N={m \choose 2}$ edges as binary (presence/absence) variables $X_i$; in this case, the length $N$ of data sequence may naturally increase as a function of $m$.) For these exponential models, the dimension of the parameter $\boldsymbol \theta_N$ is the same for each $N$ (i.e., $\boldsymbol \theta_N$ lies in a parameter space of fixed Euclidean dimension $k$).  This need not be true for other types of FSFS models considered in Sections \ref{restricted-boltzmann-machines}-\ref{deep-learning}. @schweinberger2011instability considered instability in such exponential models (e.g., for random graphs) for sequences of fixed parameters $\boldsymbol \theta_N=\boldsymbol \lambda\in\mathbb{R}^k$, $N \geq 1$, of non-varying dimension $k$.

@@

A restricted Boltzmann machine (RBM) is an undirected graphical model specified for discrete or continuous random variables, binary variables being most common [cf. @smolensky1986information]. A RBM architecture has two layers, hidden ($\mathcal{H}$) and visible ($\mathcal{V}$), with conditional independence within each layer. Let $\boldsymbol X=(X_1,\ldots,X_N)$ denote the $N$ random variables for visibles with support $\mathcal{X}^N$ and $\boldsymbol H=(H_1,\ldots,H_{N_H})$ denote the $N_H$ random variables for hiddens with support $\mathcal{X}^{N_H}$ where  $\mathcal{X} = \{-1,1\}$. For parameters $\boldsymbol \alpha \in \mathbb{R}^{N_H}$, $\boldsymbol \beta \in \mathbb{R}^N$, and $\Gamma$ as a matrix with dimension $N_H \times N$, the RBM for $\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)$ then has the joint probability mass function
$$
P_{\boldsymbol \theta} (\tilde{\boldsymbol x}) = \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^T \Gamma \boldsymbol x - \psi(\boldsymbol \theta)\right], \qquad \tilde{\boldsymbol x} = (\boldsymbol h, \boldsymbol x) \in \mathcal{X}^{N+N_H},
$$
where 
$$
\psi(\boldsymbol \theta) = \log \sum_{\tilde{\boldsymbol x} \in \mathcal{X}^{N + N_H}} \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^T \Gamma \boldsymbol x\right], \qquad \boldsymbol \theta \in \Theta,
$$ 
is the normalizing function. Let $\boldsymbol \theta_N = (\boldsymbol \alpha, \boldsymbol \beta,\Gamma) \in \Theta \subset \mathbb{R}^{q(N)}$ with $q(N) = N + N_H + N*N_H$ denote the vector of parameters for the RBM.

The probability mass function for the visible variables $X_1, \dots, X_N$ follows from marginalizing this joint specification:
$$
P_{\boldsymbol \theta} (\boldsymbol x) = \sum\limits_{\boldsymbol h \in \mathcal{X}^{N_H}} P_{\boldsymbol \theta} (\boldsymbol x, \boldsymbol h), \qquad \boldsymbol x \in \mathcal{X}^N.
$$

Note that the vector of model parameters $\boldsymbol \theta_N$, of size $q(N)$, grows in size as a function of sample dimension $N$ to accommodate the dimension of visible variables $X_1, \dots, X_N$, and one may further choose the number $N_H$ of hidden variables to change with $N$ as well; in particular, the number $N_H$ of hiddens may also potentially increase with $N$. Additionally, as $|\mathcal{X}| = 2$ and $p_{\boldsymbol \theta}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$, the RBM model specification for visibles is a FSFS model. This example also indicates that models formed by marginalizing a base FSFS model (e.g., a type of exponential family model) is again a FSFS model class.

@@

**Deep Boltzmann machine (DBM).** The DBM class of models maintains conditional independence within all layers in the model by stacking RBM models and maintaining the restriction of only allowing conditional dependence between neighboring layers. The joint probability mass function for a DBM is
$$
P_{\boldsymbol \theta} (\tilde{\boldsymbol x}) = \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol x + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)} - \psi(\boldsymbol \theta) \right], 
$$
for $\tilde{\boldsymbol x} = (\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}, \boldsymbol x) \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)} +N}$ where
$$
\psi(\boldsymbol \theta) = \log \sum\limits_{\tilde{\boldsymbol x} \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)} +N}} \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol x + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)}\right], \qquad \boldsymbol \theta \in \Theta
$$ 
is the normalizing function and parameters in the model are $\boldsymbol \beta \in \mathbb{R}^V$, $\boldsymbol \alpha^{(i)} \in \mathbb{R}^{N_{(H,i)}}$ for $i = 1, \dots, M$, along with a matrix $\Gamma^{(0)}$ of dimension $N_{(H,1)} \times N$, and matrices $\Gamma^{(i)}$ of dimension $N_{(H,i)} \times N_{(H,i+1)}$ for $i = 1, \dots, M-1$. Let $\boldsymbol \theta = (\boldsymbol \alpha^{(1)}, \dots, \boldsymbol \alpha^{(M)}, \boldsymbol \beta,\Gamma^{(0)}, \dots, \Gamma^{(M - 1)}) \in \Theta \subset \mathbb{R}^{q(N)}$ denote the combined vector of parameters with total length $q(N)= N_{(H,1)}+\cdots N_{(H,M)} + N + N_{(H,1)}*N+N_{H,2}*H_{(H,1)}+\cdots +N_{(H,M)}*H_{(H,M)-1}$.

The probability mass function for the visible random variables $X_1, \dots, X_N$ follows from this joint specification as
$$
P_{\boldsymbol \theta} (\boldsymbol x) = \sum\limits_{(\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}) \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)}}} p_{\boldsymbol \theta} (\tilde{\boldsymbol x}), \qquad \boldsymbol x \in \mathcal{X}^N
$$

@@

To define or measure instability in FSFS models, it is useful to consider the behavior of a data model sequence $P_{\theta_N}$. A relevant quantity to this end is a (scaled) extremal log-probability ratio (ELPR)
\begin{align}
\frac{1}{N} \log \left[\frac{\max\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{\min\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta}(x_1, \dots, x_N)}\right] \equiv \frac{1}{N} \text{ELPR}_N(\boldsymbol \theta_N). \label{eqn:elpr}
\end{align}

@@

\begin{defn}[S-unstable FSFS]
\label{defn:instab}
Let $\boldsymbol \theta_N \in \mathbb{R}^{q(N)}$ be a sequence of FSFS model parameters where the size of the model $q(N)$ is a function of the number of random variables $N$. A FSFS model formulation is \emph{Schweinberger-unstable} or \emph{S-unstable} if, as the number of variables increase ($N \rightarrow \infty$), 
\begin{align*}
\lim\limits_{N \rightarrow \infty} \frac{1}{N} \text{ELPR}(\boldsymbol \theta_N) \equiv \lim\limits_{N \rightarrow \infty} \frac{1}{N} \log \left[\frac{\max\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{\min\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta}(x_1, \dots, x_N)}\right] = \infty.
\end{align*}
\end{defn}
In other words, given any $C > 0$, there exists an integer $N_C > 0$ so that $\frac{1}{N}\text{ELPR}_N(\boldsymbol \theta_N) > C$ for all $N \ge N_C$.

@@

\begin{prop}
\label{prop:instab}
Let $C > 0$ and let $\text{ELPR}(\boldsymbol \theta_N)$ be as in (\ref{eqn:elpr}) for an integer $N \ge 1$. If $$\frac{1}{N}\text{ELPR}_N(\boldsymbol \theta_N) > C,$$ 
then 
$$\Delta_N(\boldsymbol \theta_N) > C.$$
\end{prop}

@@

\begin{remark}
There is a further generalization the notion of instability in Definition \ref{defn:instab} meant to address independent replications of data sequences. That is, one might consider data as $n$ independent and identically distributed replications $\boldsymbol X_1, \dots, \boldsymbol X_n$, where each $\boldsymbol X_i=(X_{i,1},\ldots,X_{i,N}) \in \mathcal{X}^N$ follows a common FSFS model with probabilities $P_{\theta_N}(\boldsymbol x)>0$,  $\boldsymbol x\in\mathcal{X}^N$ and $|\mathcal{X}|<\infty$, for $i=1,\ldots,n$.  This leads to a total of $n*N$ random variables in the joint model. However, the definition of S-unstable and Propositions \ref{prop:instab}-\ref{prop:degen} still hold for such iid replications. This is because $\left(\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n$ is the largest probability possible under the joint model for the $n$ replications while $\left(\min\limits_{\boldsymbol x \in \mathcal{X}}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n$ is the smallest probability. Thus, 
$$
\frac{\text{extremal log-probability ratio}}{\text{\# random variables in the model}} \equiv \frac{1}{n*N}\log \left[\frac{\left(\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n}{\left(\min\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n}\right] = \frac{1}{N} \log \left[\frac{\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)}{\min\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)}\right] = \frac{1}{N} \text{ELPR}(\boldsymbol \theta_N),
$$
implying that the definition of an S-unstable FSFS model sequence is invariant to the level ($n$) of independent replication. Consequently, overall model instabilities may be characterized by those of one observation from the common FSFS model.   
\end{remark}

>>>>

---
title: "A note on the instability and degeneracy of deep learning models"
author: | 
  | Andee Kaplan, Daniel Nordman, and Stephen Vardeman
  | Iowa State University
  | ajkaplan@iastate.edu, dnordman@iastate.edu, vardeman@iastate.edu
date: "November, 2016"
linestretch: 2
output: 
  pdf_document:
    keep_tex: true
    fig_caption: yes
    includes:
      in_header: latex/header.tex
    number_sections: yes
bibliography: references/refs.bib
abstract: |
  A probability model exhibits instability if small changes in a data outcome result in large, and often unanticipated, changes in probability. For correlated data structures found in several application areas, there is increasing interest in predicting/identifying instability. We consider the problem of quantifying instability for general probability models defined on sequences of observations, where each sequence of length $N$ has a finite number of possible outcomes. (A sequence of probability models results indexed by $N$ that accommodates data of expanding dimension.) Model instability is formally shown to occur when a certain log-probability ratio under such models grows faster than $N$. In this case, a one component change in the data sequence can shift probability by orders of magnitude. Also, as a measure of instability becomes more extreme, the resulting probability models are shown  to tend to degeneracy, placing all their probability on arbitrarily small portions of the sample space. These results on instability apply to large classes of models commonly used in random graphs, network analysis, and machine learning contexts.
---

@@

Section \ref{examples} provides several examples of FSFS models commonly used in graph/network analysis and machine learning (i.e., deep learning models). Section \ref{instability-results} establishes formal results regarding the propriety of FSFS models with regard to stability. A FSFS probability model sequence exhibits instability if small changes in the components of a data outcome $(x_1,\ldots,x_N)$ can result in large changes in probability $P_{\theta_N}(x_1,\ldots,x_N)$. The concept of instability, introduced in the field of statistical physics by @ruelle1999statistical, was extended to include a notion of detection and quantification for certain exponential 
family models by @schweinberger2011instability. For similar exponential models, particularly in connection to random graphs/networks, @handcock2003assessing considered (mean-based) characterizations for so-called model degeneracy, whereby a probability model places all mass on a small subset of the sample space and produces undesirably low variability in model outcomes. As described by @schweinberger2011instability, model instability and model degeneracy are related by viewing degeneracy as an extreme or limiting form of instability. The instability results of @schweinberger2011instability were developed for the case of discrete exponential family models. The main results here concern a general measure of model instability, appropriate across the whole FSFS model class. This can be used to identify when certain maximal probabilities in FSFS models are too extreme relative to the length $N$. In this case, a one component change in the data sequence may shift probability by orders of magnitude, and FSFS models are rigorously shown to become degenerate as the measure of instability increases. Lastly, Section \ref{implications} emphasizes the implications of our model propriety results and proofs of the main results appear in the Appendix.

@@

Defining $P_{\boldsymbol \theta_N}(\boldsymbol x)\equiv p_{N,\boldsymbol \lambda_N}(\boldsymbol x)$ with $\boldsymbol \theta_N=\boldsymbol \lambda_N$ to be a sequence of elements of $\Lambda \subset \mathbb{R}^k$ and noting that $P_{\boldsymbol \theta_N}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$, these discrete exponential family models are special cases of the FSFS models. Such exponential models arise with spatial data on a lattice [@besag1974spatial], network data [@wasserman1994social; @handcock2003assessing], and even binomial sampling with $N$ iid Bernoulli random variables. (Note that for random graphs or networks with, say, $m$ nodes, one may wish to consider $N={m \choose 2}$ edges as binary (presence/absence) variables $X_i$. In this case, the length $N$ of data sequence may naturally increase as a function of $m$.) For these exponential models, the dimension of the parameter $\boldsymbol \theta_N$ is the same for each $N$ as $\boldsymbol \theta_N$ lies in a parameter space of fixed Euclidean dimension $k$.  This need not be true for other types of FSFS models considered in Sections \ref{restricted-boltzmann-machines}-\ref{deep-learning}. @schweinberger2011instability considered instability in such exponential models (e.g., for random graphs) for sequences of fixed parameters $\boldsymbol \theta_N=\boldsymbol \lambda\in\mathbb{R}^k$, $N \geq 1$, of non-varying dimension $k$.

@@

A restricted Boltzmann machine (RBM) is an undirected graphical model specified for discrete or continuous random variables, binary variables being most common [cf. @smolensky1986information]. A RBM architecture has two layers, hidden ($\mathcal{H}$) and visible ($\mathcal{V}$), with conditional independence within each layer. Let $\boldsymbol X=(X_1,\ldots,X_N)$ denote the $N$ random variables for visibles with support $\mathcal{X}^N$ and $\boldsymbol H=(H_1,\ldots,H_{N_H})$ denote the $N_H$ random variables for hiddens with support $\mathcal{X}^{N_H}$ where  $\mathcal{X} = \{-1,1\}$. For parameters $\boldsymbol \alpha \in \mathbb{R}^{N_H}$, $\boldsymbol \beta \in \mathbb{R}^N$, and $\Gamma$ as a matrix with dimension $N_H \times N$, the RBM for $\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)$ then has the joint probability mass function
$$
P_{\boldsymbol \theta_N} (\tilde{\boldsymbol x}) = \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^T \Gamma \boldsymbol x - \psi(\boldsymbol \theta_N)\right], \qquad \tilde{\boldsymbol x} = (\boldsymbol h, \boldsymbol x) \in \mathcal{X}^{N+N_H},
$$
where 
$$
\psi(\boldsymbol \theta_N) = \log \sum_{\tilde{\boldsymbol x} \in \mathcal{X}^{N + N_H}} \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^T \Gamma \boldsymbol x\right], \qquad \boldsymbol \theta_N \in \Theta_N,
$$ 
is the normalizing function. Let $\boldsymbol \theta_N = (\boldsymbol \alpha, \boldsymbol \beta,\Gamma) \in \Theta_N \subset \mathbb{R}^{q(N)}$ with $q(N) = N + N_H + N*N_H$ denote the vector of parameters for the RBM.

The probability mass function for the visible variables $X_1, \dots, X_N$ follows from marginalizing this joint specification:
$$
P_{\boldsymbol \theta_N} (\boldsymbol x) = \sum\limits_{\boldsymbol h \in \mathcal{X}^{N_H}} P_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h), \qquad \boldsymbol x \in \mathcal{X}^N.
$$

Note that the vector of model parameters $\boldsymbol \theta_N$, of size $q(N)$, grows in size as a function of sample dimension $N$ to accommodate the dimension of visible variables $X_1, \dots, X_N$, and one may further choose the number $N_H$ of hidden variables to change with $N$ as well; in particular, the number $N_H$ of hiddens may also potentially increase with $N$. Additionally, as $|\mathcal{X}| = 2$ and $P_{\boldsymbol \theta_N}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$, the RBM model specification for visibles is a FSFS model. This example also indicates that models formed by marginalizing a base FSFS model (e.g., a type of exponential family model) is again a FSFS model class.

@@

**Deep Boltzmann machine (DBM).** The DBM class of models maintains conditional independence within all layers in the model by stacking RBM models and maintaining the restriction of only allowing conditional dependence between neighboring layers. The joint probability mass function for a DBM is
$$
P_{\boldsymbol \theta_N} (\tilde{\boldsymbol x}) = \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol x + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)} - \psi(\boldsymbol \theta_N) \right], 
$$
for $\tilde{\boldsymbol x} = (\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}, \boldsymbol x) \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)} +N}$ where
$$
\psi(\boldsymbol \theta_N) = \log \sum\limits_{\tilde{\boldsymbol x} \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)} +N}} \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol x + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)}\right], \qquad \boldsymbol \theta_N \in \Theta_N
$$ 
is the normalizing function and parameters in the model are $\boldsymbol \beta \in \mathbb{R}^V$, $\boldsymbol \alpha^{(i)} \in \mathbb{R}^{N_{(H,i)}}$ for $i = 1, \dots, M$, along with a matrix $\Gamma^{(0)}$ of dimension $N_{(H,1)} \times N$, and matrices $\Gamma^{(i)}$ of dimension $N_{(H,i)} \times N_{(H,i+1)}$ for $i = 1, \dots, M-1$. Let $\boldsymbol \theta_N = (\boldsymbol \alpha^{(1)}, \dots, \boldsymbol \alpha^{(M)}, \boldsymbol \beta,\Gamma^{(0)}, \dots, \Gamma^{(M - 1)}) \in \Theta_N \subset \mathbb{R}^{q(N)}$ denote the combined vector of parameters with total length $q(N)= N_{(H,1)}+\cdots N_{(H,M)} + N + N_{(H,1)}*N+N_{H,2}*H_{(H,1)}+\cdots +N_{(H,M)}*H_{(H,M)-1}$.

The probability mass function for the visible random variables $X_1, \dots, X_N$ follows from this joint specification as
$$
P_{\boldsymbol \theta_N} (\boldsymbol x) = \sum\limits_{(\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}) \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)}}} P_{\boldsymbol \theta_N} (\tilde{\boldsymbol x}), \qquad \boldsymbol x \in \mathcal{X}^N
$$

@@

To define or measure instability in FSFS models, it is useful to consider the behavior of a data model sequence $P_{\theta_N}$. A relevant quantity to this end is a (scaled) extremal log-probability ratio (ELPR)
\begin{align}
\frac{1}{N} \log \left[\frac{\max\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{\min\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}\right] \equiv \frac{1}{N} \text{ELPR}_N(\boldsymbol \theta_N). \label{eqn:elpr}
\end{align}

@@

\begin{defn}[S-unstable FSFS]
\label{defn:instab}
Let $\boldsymbol \theta_N \in \mathbb{R}^{q(N)}$ be a sequence of FSFS model parameters where the size of the model $q(N)$ is a function of the number of random variables $N$. A FSFS model formulation is \emph{Schweinberger-unstable} or \emph{S-unstable} if, as the number of variables increase ($N \rightarrow \infty$), 
\begin{align*}
\lim\limits_{N \rightarrow \infty} \frac{1}{N} \text{ELPR}(\boldsymbol \theta_N) \equiv \lim\limits_{N \rightarrow \infty} \frac{1}{N} \log \left[\frac{\max\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{\min\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}\right] = \infty.
\end{align*}
\end{defn}
In other words, given any $C > 0$, there exists an integer $N_C > 0$ so that $\frac{1}{N}\text{ELPR}_N(\boldsymbol \theta_N) > C$ for all $N \ge N_C$.

@@

\begin{prop}
\label{prop:instab}
Let $\text{ELPR}(\boldsymbol \theta_N)$ be as in (\ref{eqn:elpr}) for an integer $N \ge 1$. For a given $C>0$, if $$\frac{1}{N}\text{ELPR}_N(\boldsymbol \theta_N) > C,$$ 
then 
$$\Delta_N(\boldsymbol \theta_N) > C.$$
\end{prop}

@@

\begin{remark}
There is a further generalization the notion of instability in Definition \ref{defn:instab} meant to address independent replications of data sequences. That is, one might consider data as $n$ independent and identically distributed replications $\boldsymbol X_1, \dots, \boldsymbol X_n$, where each $\boldsymbol X_i=(X_{i,1},\ldots,X_{i,N}) \in \mathcal{X}^N$ follows a common FSFS model with probabilities $P_{\theta_N}(\boldsymbol x)>0$,  $\boldsymbol x\in\mathcal{X}^N$ and $|\mathcal{X}|<\infty$, for $i=1,\ldots,n$.  This leads to a total of $n*N$ random variables in the joint model. However, the definition of S-unstable and Propositions \ref{prop:instab}-\ref{prop:degen} still hold for such iid replications. This is because $\left(\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n$ is the largest probability possible under the joint model for the $n$ replications while $\left(\min\limits_{\boldsymbol x \in \mathcal{X}}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n$ is the smallest probability. Thus, for the combined replications $\boldsymbol X_1, \ldots, \boldsymbol X_n$, the analog definition of the extremal log-probability becomes
$$
\frac{\text{extremal log-probability ratio}}{\text{\# random variables in the model}} \equiv \frac{1}{n*N}\log \left[\frac{\left(\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n}{\left(\min\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n}\right] = \frac{1}{N} \log \left[\frac{\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)}{\min\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)}\right] = \frac{1}{N} \text{ELPR}(\boldsymbol \theta_N),
$$
implying that the definition of an S-unstable FSFS model sequence is invariant to the level ($n$) of independent replication. Consequently, overall model instabilities may be characterized by those of one observation from the common FSFS model.   
\end{remark}
